---
title: "Acknowledgments and resources"
# general Output Options
execute:     
  eval: true
  echo: false     #  include source code in output?
  warning: false  #  include warning code in output?
  error: false    #  include error code in output?
  output: true   # include output code in output (CHG in BLOCKS)?
  # include: false   # R still runs but code and results DON"T appear in output  
  # cache: false
toc: true
format:
  html:
    code-fold: false # redundant IF echo false 
    toc-depth: 2
    toc_float: true
    toc-location: left
    toc-title: ""
    # template: theme/template.tex  
bibliography: bib/R4biostats.bib
csl: bib/apa-cv.csl # Very important for complete citations!!!!! 
## --------- NO END REFERENCES .... other options are redundant!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!
suppress-bibliography: true
# biblio-style: apa # authoryear-icomp # abbrv # alpha #apalike # plain # unsrt #siam
#reference-section-title: References
#link-citations: false
# filters:
#   - section-bibliographies
---

<!-- ::: {.callout-caution collapse="false"} -->
<!-- Website construction in progress... -->
<!-- ::: -->

```{r}
if (!require(pacman)) {install.packages(pacman)} 
library(pacman)

# pacman::p_ver("bib2df")
# pacman::p_exists("bib2df")  #TRUE
# pacman::p_ver_diff("bib2df")
# p_load_gh("ropensci/bib2df") # otherwise warns `as_data_frame() was deprecated in tibble 2.0.0.`

#remotes::install_github("ropensci/bib2df")
p_load(RefManageR,
       #  bib2df,
       bibtex,
       janitor,here, paint,
       glue,
       tidyverse, # lubridate,
       htmltools,
       # distilltools,
       bib2df
)

```

```{r}
# For this I need the Zotero bib to be exported as BEtter BibLaTex !!!!!!!!!!!!!

# - Load bibliography (exported from Zotero)
bib_path <- here::here ("bib","R4biostats.bib")

# - as list  (NEEDED for functions later)
mybib <- RefManageR::ReadBib(bib_path, check = FALSE,.Encoding = "UTF-8" )  #bibtex::do_read_bib(file = bib_path, encoding = "unknown", srcfile)

# - as dataframe
mybib_df_raw <- bib2df::bib2df(bib_path)  

# - Format dataframe version of bibliography
mybib_df <-
  mybib_df_raw %>%
  clean_names() %>%
  select(key = bibtexkey, 
         date, 
         author, 
         title, 
         category,# THIS MY Zotero's 
         type, # THIS MY OWN "Extra: tex.type: ..." in zotero 
         url, 
         file, 
         include # THIS MY OWN "Extra: tex.include: ..." in zotero 
         ) %>%
  mutate(
    # date = parse_date_time(date, "ymd",
    # 							  #select_formats = .select_formats
    # 							  drop = FALSE),
    yyyy_mm = str_sub(date, start = 1, end = 7), # I have some irregularities in format  
    date_parsed = parse_date_time(yyyy_mm,  "ym", truncated = 2 ),
    year = year(date_parsed), 
    month = month(date_parsed)) %>%
  arrange(desc(year), key) %>% 
  dplyr::relocate(c("yyyy_mm" ,    "year" ,        "month"  ), .after =date)

   
# # Set RefManageR options (needed to avoid the "[1]")
# oldopts <- RefManageR::BibOptions(
# 	style = "text",
# 	bib.style = "authoryear", 
# 	no.print.fields = c("doi", "url", "urldate", "eprint", "eprinttype"),
# 	max.names = 50, # maximum number of names to display before using “et al.”
# 	dashed = FALSE, 
# 	sorting = "ydnt"
# )

# # - Load additional metadata
# mybib_meta <- tibble::tribble( 
# 	~"url_pub", ~"url_pdf",~ "url_rg", ~"url_repo", ~"url_arch", ~"url_osf", ~"url_other", ~"url_blog" ) 
# mybib_meta[1:nrow(mybib_df), ] <- as.character(NA)
# 
# # bind 
# mybib_df2 <- cbind(mybib_df, mybib_meta)
# names(mybib_df2)
# rm(mybib, mybib_df, mybib_df_raw, mybib_meta)
```

 
<!-- ## Acknowledgements -->
<!-- something along the lines of [MADA course](https://andreashandel.github.io/MADAcourse/)    -->
<!--  https://www.laleggepertutti.it/492987_fair-use-esiste-in-italia -->

We are genuinely grateful to many people within the Statistics, Machine Learning, Epidemiology and R programming communities, who shared their valuable work, open source software, and training resources. <!-- Special thanks to [Luisa M. Mimmi]({{< var websites.lulas_site >}}) (my sister!) who revised the workshop content and built this dedicated website.  -->

<!-- The following individuals have contributed to improving the course: Ian Bachli (@idbachli), Megan Beaudry (@MeganSarah), Zane Billings (@wzbillings),   -->

<!-- I would love to hear your feedback, questions, or suggestions about the workshop:  <p> <a id="mail-link" href="mailto:lmm76@georgetown.edu"  class="btn btn-info"> <i class="fa fa-envelope"></i> &nbsp;&nbsp;contact us</a></p> -->

## Licensing and use of the workshop materials

The workshop materials are licensed under the [Creative Commons Attribution-NonCommercial-ShareAlike 4.0 International License](https://creativecommons.org/licenses/by-nc-sa/4.0/). All borrowed external materials (images, worked examples, etc.), are credited with proper "*Source*" statements and governed by their own licenses. To our knowledge, all the consulted materials were published under "open access" or "creative commons" frameworks. If this were not the case for any content piece displayed here, please let us know and it will be removed.

 
<!-- ## Materials -->

<!-- This source code and license for all of the materials on this website can be found on a dedicated Github repo [{{< fa brands github >}}](https://github.com/Lulliter/R4biostats). -->

<!-- ## License -->

<!-- The work presented in this website is licensed under a {{< fa brands creative-commons >}} {{< fa brands creative-commons-by >}} {{< fa brands creative-commons-nc >}} {{< fa brands creative-commons-sa >}} [Creative Commons Attribution-NonCommercial-ShareAlike 4.0 International License](https://creativecommons.org/licenses/by-nc-sa/4.0/), except for the borrowed and mentioned with proper *"Source:"* statements. -->


## Selected resources for self-guided learning

Below is a curated list of great resources (most of which free and openly accessible) you can peruse on your own. 


<!-- ### Journal Articles -->

```{r article}
#| include: false

#mybib_df[mybib_df$type == "article",  1] 
mybib_df %>% 
  filter(!(include %in% "no") & type ==  "article") %>% 
   select(key)

# print(mybib[type = "article"], 
#       .opts = list(bib.style = "authoryear", 
#                    check.entries = FALSE))

```

 <!-- + @mayne_role_1957     -->

### Biostatistics/Epidemiology with R examples

```{r course}
#| include: false

#mybib_df[mybib_df$type == "course",  1] 
mybib_df %>% 
  filter(!(include %in% "no") & type == "course") %>% 
  select(key) %>% 
  # just to copy by hand
  map( ~cat (paste0("+ @", .x),
             fill = FALSE, 
             labels = NULL, sep = "\n")
  )
```

```{r book}
#| include: false

#mybib_df[mybib_df$type == "course",  1] 
mybib_df %>% 
  filter(!(include %in% "no") & type == "book") %>% 
  select(key) %>% 
  # just to copy by hand
  map( ~cat (paste0("+ @", .x),
             fill = FALSE, 
             labels = NULL, sep = "\n")
  )
```


+ @applied_epi_team_applied_2024
+ @applied_epi_team_r_2024
+ @bobbitt_statology_2024
+ @cetinkaya-rundel_introduction_2023
+ @childs_introductory_2022-1
+ @selby_analytical_2024                                   
+ @vu_introductory_2021

### R packages & tools

```{r R}
#| include: false

#mybib_df[mybib_df$type == "R",  1] 
mybib_df %>% 
  filter(!(include %in% "no") & type == "R") %>% 
   select(key) %>% 
  # just to copy by hand
  map( ~cat (paste0("+ @", .x),
             fill = FALSE, 
             labels = NULL, sep = "\n")
  )
``` 

+ @posit_posit_2024
+ @rstudio_base-r-cheat-sheetpdf_2024
+ @various_contributors_bioconductor_2024
+ @wickham_programming_2023
+ @wickham_tidy_2014


### Sources of practice datasets

```{r dataset}
#| include: false
mybib_df %>% 
  filter(!(include %in% "no") & type == "dataset") %>% 
   select(key)  %>% 
  # just to copy by hand
  map( ~cat (paste0("+ @", .x),
             fill = FALSE, 
             labels = NULL, sep = "\n")
  )
```

+ @vanderbilt_department_of_biostatistics_vanderbilt_2023
+ @chicco_machine_2020
+ @ahmad_survival_2017

<!-- <!-- For HTML output -->  
<!-- ::: {.content-hidden when-format="html"} -->
<!-- ## References -->
<!-- ::: -->

