---
title: "Diving deeper into Causal Analysis"
subtitle: "<span style='font-size:2em;'> Practice session covering topics ..... </span>"
author: "<a href=\"{{< var websites.lulas_site >}}\" style='color:#72aed8;font-weight:600;'>Luisa M. Mimmi, M.P.P. </a>&ensp;|&ensp; Independent consultant"
date: 2024-08-27
date-format: long
code-link: true
format:
  revealjs:
    smaller: true
    scrollable: true
    theme: ../../theme/slidesMine.scss # QUARTO LOOKS IN SAME FOLDER 
#    logo: imgs_slides/mitgest_logo.png
    footer: '[{{< var websites.live_site >}}]("{{< var websites.live_site >}}") | R 4 stats'
#    footer: <https://lulliter.github.io/R4biostats/lectures.html>
## ------------- x salvare come PDF 
    standalone: false
    ## -------Produce a standalone HTML file with no external dependencies,
    embed-resources: true
    transition: fade
    background-transition: fade
    highlight-style: ayu-mirage
    slide-number: true
    fig-cap-location: top
    # fig-format: svg
    pdf-separate-fragments: false
    # fig-align: center
execute:
  # Quarto pre code blocks do not echo their source code by default
  echo: true
  include: true
  freeze: auto
bibliography: ../../bib/R4biostats.bib
csl: ../../bib/apa-6th-edition.csl 
suppress-bibliography: true
---

# [GOAL OF TODAY'S PRACTICE SESSION]{.r-fit-text}

::: {style="font-size: 100%;"}
::: {style="color:#77501a"}
+ ......
:::
:::

<br><br>

::: {style="font-size: 70%;"}
The examples and datasets in this Lab session follow very closely two sources:

1. ....
2. ....
<!-- https://mvanrongen.github.io/corestats-in-r_tidyverse/power-analysis.html -->
:::


# R ENVIRONMENT SET UP & DATA

## [Needed R Packages]{.r-fit-text}

::: {style="font-size: 85%;"}

+ We will use functions from packages `base`, `utils`, and `stats` (pre-installed and pre-loaded) 
+ We may also use the packages below (specifying `package::function` for clarity).


```{r}
# Load pckgs for this R session

# --- General 
library(here)     # tools find your project's files, based on working directory
library(dplyr)    # A Grammar of Data Manipulation
library(skimr)    # Compact and Flexible Summaries of Data
library(magrittr) # A Forward-Pipe Operator for R 
library(readr)    # A Forward-Pipe Operator for R 

# Plotting & data visualization
library(ggplot2)      # Create Elegant Data Visualisations Using the Grammar of Graphics
library(ggfortify)     # Data Visualization Tools for Statistical Analysis Results
library(scatterplot3d) # 3D Scatter Plot

# --- Statistics
library(MASS)       # Support Functions and Datasets for Venables and Ripley's MASS
library(factoextra) # Extract and Visualize the Results of Multivariate Data Analyses
library(FactoMineR) # Multivariate Exploratory Data Analysis and Data Mining
library(rstatix)    # Pipe-Friendly Framework for Basic Statistical Tests

# --- Tidymodels (meta package)
library(rsample)    # General Resampling Infrastructure  
library(broom)      # Convert Statistical Objects into Tidy Tibbles
```

<!-- # Data  -->
<!-- # devtools::install_github("OI-Biostat/oi_biostat_data") -->
<!-- #library(oibiostat) # Data Package for OpenIntro Biostatistics  -->
:::



# TO DO 

RIMESCOLARE LA LEZ 3 CON QUESTO https://gabors-data-analysis.com/images/slides-public/da-public-slides-ch19-v3-2023.pdf

# DATASETS for today

<br>

::: {style="font-size: 80%;"}
In this tutorial, we will use: 

+ .......

:::


## [Dataset on ....]{.r-fit-text}

::: {style="font-size: 95%;"}
**Name**: ....
**Documentation**: ....
**Sampling details**: ....
::: 

## [Importing Dataset `....`]{.r-fit-text}

.

+ The data can be interactively obtained form the `MASS` R package

```{r}
# (after loading pckg)
# library(MASS)  

# I can call 
# utils::data(biopsy)
```


## []{.r-fit-text}

We define:

+ the **individual treatment effect**
+ the **average treatment effect**
+ and **the average treatment effect on the treated**

## []{.r-fit-text}


## []{.r-fit-text}

## []{.r-fit-text}


## []{.r-fit-text}


## []{.r-fit-text}


## []{.r-fit-text}





 
## Final thoughts

::: {style="font-size: 95%;"}
::: {style="color:#77501a"}


+ .....
  + ...
+ ...


:::

:::

