---
title: "Lab 4:Mapping causal & predictive approaches"
subtitle: "<span style='font-size:2em;'> Practice session</span>"
author: "<a href='https://lulliter.github.io/R4stats/' style='color:#72aed8;font-weight:600;'>Luisa M. Mimmi</a>&ensp;|&ensp;[https://luisamimmi.org/](https://luisamimmi.org/)"
date: 2024-07-27
date-format: long
code-link: true
format:
  revealjs:
    smaller: true
    scrollable: true
    theme: ../../theme/slidesMine.scss # QUARTO LOOKS IN SAME FOLDER 
#    logo: imgs_slides/mitgest_logo.png
    footer: '[R 4 Statistics](https://lulliter.github.io/R4stats/) | 2024'
#    footer: <https://lulliter.github.io/R4biostats/lectures.html>
## ------------- x salvare come PDF 
    standalone: false
    ## -------Produce a standalone HTML file with no external dependencies,
    embed-resources: true
    transition: fade
    background-transition: fade
    highlight-style: ayu-mirage
    slide-number: true
    fig-cap-location: top
    # fig-format: svg
    pdf-separate-fragments: false
    # fig-align: center
execute:
  # Quarto pre code blocks do not echo their source code by default
  echo: true
  include: true
  freeze: auto
bibliography: ../../bib/R4biostats.bib
csl: ../../bib/apa-6th-edition.csl 
suppress-bibliography: true
---


```{r}
#| echo: FALSE

# set data dir, data used   -----------------------------------------------------------
data_in <- fs::path(here::here("practice","data_input", "04_datasets", "Gabor"))

# load theme and functions
#source( here::here(data_in ,"helper.R"))
options(digits = 3)


# Functions for saving plots -----------------------------------------------------------
f_save_plot <- function(plot_name, plot_object) {
   # Print the plot, save as PDF and PNG
   plot_object %T>%
      print() %T>%
      ggsave(., filename = here("practice", "practice_slides", "figures", paste0(plot_name, ".pdf")),
             # width = 4, height = 2.25, units = "in",
             device = cairo_pdf) %>%
      ggsave(., filename = here("practice", "practice_slides", "figures", paste0(plot_name, ".png")),
             # width = 4, height = 2.25, units = "in",
             type = "cairo", dpi = 300)
}

f_descr_table <- function(var) {
  workfile %>%
    summarise(mean=mean(var),
              median=median(var),
              std_dev=sd(var),
              minimum=min(var),
              maximum=max(var), 
              observations=n())
}
# Example of using the function
# f_save_plot("proj_wrd_freq", proj_wrd_freq)

```

# [GOAL OF TODAY'S PRACTICE SESSION]{.r-fit-text}

::: {style="font-size: 100%;"}
::: {style="color:#77501a"}
+ ...
:::
:::

<br><br>

::: {style="font-size: 70%;"}
The examples and datasets in this Lab session follow very closely two sources:

1. ...
:::


## Topics discussed in Lecture # 4


::: {style="font-size: 95%;"}
**Lecture 4: topics** 
<!-- + Shifting the emphasis on **empirical prediction**  -->
<!--   + Distinction between supervised & **unsupervised** algorithms -->
<!--     + Unsupervised ML Example -->
<!--       + PCA  -->

+ ....

<!-- + Workshop Conclusions -->

:::  

# R ENVIRONMENT SET UP & DATA

## Needed R Packages
::: {style="font-size: 85%;"}

+ We will use functions from packages `base`, `utils`, and `stats` (pre-installed and pre-loaded) 
+ We may also use the packages below (specifying `package::function` for clarity).


```{r}
# Load pckgs for this R session

# --- General 
library(here) # A Simpler Way to Find Your Files   
library(skimr)    # Compact and Flexible Summaries of Data
library(paint)   # Fancy structure for data frames 
library(janitor)  # Simple Tools for Examining and Cleaning Dirty Data
library(tidyverse) # Easily Install and Load the 'Tidyverse'

# --- Plotting & data visualization
library(ggfortify)     # Data Visualization Tools for Statistical Analysis Results
library(ggpubr)        # 'ggplot2' Based Publication Ready Plots
library(dagitty) # Graphical Analysis of Structural Causal Models
library(ggdag) # Analyze and Create Elegant Directed Acyclic Graphs #
library(patchwork) # The Composer of Plots

# --- Descriptive statistics and regressions
library(broom) # Convert Statistical Objects into Tidy Tibbles
library(Hmisc) # Harrell Miscellaneous
library(estimatr) # Fast Estimators for Design-Based Inference
library(modelsummary) # Summary Tables and Plots for Statistical Models and Data:
library(sandwich) #for robust variance estimation
library(survey) # Analysis of Complex Survey Samples
library(haven) # Import and Export 'SPSS', 'Stata' and 'SAS' Files
library(simstudy) # Simulation of Study Data
library(NHANES) # Data from the US National Health and Nutrition Examination Study  
```
:::


<!-- ## DAG to represent confounders -->

<!-- ```{r} -->
<!-- #| output-location: slide -->

<!-- dag <- dagitty::dagitty("dag { -->
<!--   X -> Y -->
<!--   Z -> X -->
<!--   Z -> Y -->
<!--   e -> Y -->
<!-- }") -->

<!-- dagitty::coordinates(dag ) <-  list( -->
<!--   x=c(X=1, Z=3, Y=5, e=5), -->
<!--   y=c(X=1, Z=3, Y=1, e=2)) -->

<!-- library(ggdag) # Analyze and Create Elegant Directed Acyclic Graphs # Analyze and Create Elegant Directed Acyclic Graphs # Analyze and Create Elegant Directed Acyclic Graphs # Analyze and Create Elegant Directed Acyclic Graphs # Analyze and Create Elegant Directed Acyclic Graphs # Analyze and Create Elegant Directed Acyclic Graphs # Analyze and Create Elegant Directed Acyclic Graphs -->
<!-- dag <-  tidy_dagitty(dag) -->
<!-- ex1 <-  ggdag(dag, layout = "circle") +  -->
<!--   theme_dag_blank(plot.caption = element_text(hjust = 1)) +  -->
<!--   geom_dag_node(color="pink") +  -->
<!--   geom_dag_text(color="white") + -->
<!--   ggtitle("a) DAG with a confounder (Z)") + -->
<!--   labs(caption = "X = drug exposure\nY = outcome\nZ = confounder\n e = random error") -->
<!-- ex1 -->
<!-- ``` -->


# --- 

## ITE 

Individual Treatment Effect (ITE) is defined as the difference in potential outcomes for a given individual 
$ITE_i= y_{i}^1  − y_{i}^0$

The Causal Effect (CE)
 
$CE_i = Y_{1i} - Y_{0i}$ where $Y_{1i}$ and $Y_{0i}$ are the potential outcomes for each individual $i$ being *both* exposed and not exposed respectively (not possible in the real world).

Simulation 

```{r}
def <- simstudy::defData(varname = "C", formula = 0.4, dist = "binary")
def <- simstudy::defData(def, "X", formula = "0.3 + 0.4 * C", dist = "binary")
def <- simstudy::defData(def, "e", formula = 0, variance = 2, dist = "normal")
def <- simstudy::defData(def, "Y0", formula = "2 * C + e", dist="nonrandom")
def <- simstudy::defData(def, "Y1", formula = "1 + 2 * C + e", dist="nonrandom")
def <- simstudy::defData(def, "Y_obs", formula = "Y0 + (Y1 - Y0) * X", dist = "nonrandom") #  = Y1*X + (1-X)*Y0

set.seed(2017)
dt <- genData(10000, def)
```

```{r}
paste0("Calculated mean causal effect = ", mean(dt[, Y1] - dt[, Y0])) #mean difference of counterfactual outcomes
paste0("Calculated observed difference  = ", round(dt[X == 1, mean(Y_obs)] - dt[X == 0, mean(Y_obs)],2))

lm1 <- lm(Y_obs ~ X, data = dt)
tidy(lm1)
```

## DAG with COLLIDER

```{r}
#| output-location: slide

# df DAG 
dag_collid <- dagify(
  Y ~ X + e,
  Z ~ X + Y,
  exposure = "X",
  outcome = "Y",
  #latent = "e",
  # labels
  labels = c(
    "Z" = "Collider",
    "X" = "Exposure",
    "Y" = "Outcome",
    "e" = "Unobserved \nerror"),
  # positions
  coords = list(
    x = c(X = 0, Y= 4, Z = 2 , e = 5),
    y = c(X = 0, Y = 0, Z = 2, e = 1) 
  )) %>% 
  tidy_dagitty() 

# Plot DAG 
dag_col_p <-  dag_collid %>% 
  ggdag(., layout = "circle") +  # decided in dagify
  theme_dag_blank(plot.caption = element_text(hjust = 1)) +
  # Nodes 
  geom_dag_node(aes(color = name)) +  
  # Labels dodged to avoid overlap
  geom_dag_label(aes(label = label), color = "#4c4c4c", nudge_x = 0.7, nudge_y = 0.2) +  
  
  geom_dag_text(color="white") +
  labs(title = "Causal map with COLLIDER (Z)" , 
       #subtitle = "X = exposure, Y = outcome, Z = collider, e = random error", 
       caption = ) +
  # Map colors to specific nodes
  scale_color_manual(values = c("X" = "#9b6723", "Y" = "#72aed8", "Z" = "#d02e4c",
                                "e" = "#A6A6A6"), 
                     guide = "none")    # Remove legend

dag_col_p
```


```{r}
#| echo: FALSE
f_save_plot("dag_col_p", dag_col_p)
```


## DAG with CONFOUNDER  

```{r}
#| output-location: slide
library(dagitty) # Graphical Analysis of Structural Causal Models # Graphical Analysis of Structural Causal Models # Graphical Analysis of Structural Causal Models 
library(ggdag) # Analyze and Create Elegant Directed Acyclic Graphs # Analyze and Create Elegant Directed Acyclic Graphs # Analyze and Create Elegant Directed Acyclic Graphs 

# df DAG 
dag_confounder <- dagify(
  Y ~ X + Z + e,
  X ~  Z,
  exposure = "X",
  outcome = "Y",
  #latent = "e",
  # labels
  labels = c(
    "Z" = "Confounder",
    "X" = "Exposure",
    "Y" = "Outcome",
    "e" = "Unobserved \nerror"),
  # positions
  coords = list(
    x = c(X = 0, Y= 4, Z = 2 , e = 5),
    y = c(X = 0, Y = 0, Z = 2, e = 1) 
  )) %>% 
  tidy_dagitty()     # Add edge_type within pipe

# Plot DAG 
dag_conf_p <-  dag_confounder %>% 
  ggdag(., layout = "circle") + 
  theme_dag_blank(plot.caption = element_text(hjust = 1)) +
  # Nodes 
  geom_dag_node(aes(color = name)) +  
  # Labels dodged to avoid overlap
  geom_dag_label(aes(label = label), color = "#4c4c4c", nudge_x = 0.7, nudge_y = 0.2) +  
  geom_dag_text(color="white") +
  labs(title = "Causal map with CONFOUNDER (Z)" , #subtitle = " ",
        caption = ) +
  # Map colors to specific nodes
  scale_color_manual(values = c("X" = "#9b6723", "Y" = "#72aed8", "Z" = "#d02e4c",
                                "e" = "#A6A6A6"), 
                     guide = "none")   

dag_conf_p
```


```{r}
#| echo: FALSE
f_save_plot("dag_conf_p", dag_conf_p)
```

## DAG with MEDIATOR

```{r}
#| output-location: slide
library(dagitty) # Graphical Analysis of Structural Causal Models # Graphical Analysis of Structural Causal Models # Graphical Analysis of Structural Causal Models 
library(ggdag) # Analyze and Create Elegant Directed Acyclic Graphs # Analyze and Create Elegant Directed Acyclic Graphs # Analyze and Create Elegant Directed Acyclic Graphs 

# df DAG 
dag_med <- dagify(
  Y ~ Z + e,
  Z ~ X,
  
  exposure = "X",
  outcome = "Y",
  #latent = "e",
  # labels
  labels = c(
    "Z" = "Mediator",
    "X" = "Exposure",
    "Y" = "Outcome",
    "e" = "Unobserved \nerror"),
  # positions
  coords = list(
    x = c(X = 0, Y= 4, Z = 2 , e = 5),
    y = c(X = 0, Y = 0, Z = 2, e = 1) 
  )) %>% 
  tidy_dagitty()     # Add edge_type within pipe

# Plot DAG 
dag_med_p <-  dag_med %>% 
  ggdag(., layout = "circle") +  # decided in dagify
  theme_dag_blank(plot.caption = element_text(hjust = 1)) +
  # Nodes 
  geom_dag_node(aes(color = name)) +  
  # Labels dodged to avoid overlap
  geom_dag_label(aes(label = label), color = "#4c4c4c", nudge_x = 0.7, nudge_y = 0.2) +  
  
  geom_dag_text(color="white") +
  labs(title = "Causal map with MEDIATOR (Z)" , 
       #subtitle = "X = exposure, Y = outcome, Z = collider, e = random error", 
       caption = ) +
  # Map colors to specific nodes
  scale_color_manual(values = c("X" = "#9b6723", "Y" = "#72aed8", "Z" = "#d02e4c",
                                "e" = "#A6A6A6"), 
                     guide = "none")    # Remove legend

dag_med_p
```


```{r}
#| echo: FALSE
f_save_plot("dag_med_p", dag_med_p)
```

# ----

# DATASETS for today

<!-- <br> -->

<!-- ::: {style="font-size: 80%;"} -->
<!-- In this tutorial, we will use:  -->

<!-- + .... -->

<!-- ::: -->


## [Importing Dataset 1 (NHANES)]{.r-fit-text}

::: {style="font-size: 95%;"}
**Name**: NHANES, accessible from package `NHANES`
**Documentation**: See reference on the data downloaded and ....
**Sampling details**: We’ll use a subset of the NHANES dataset, focusing on variables related to smoking status (SmokeNow), systolic blood pressure (BPSysAve), Body Mass Index (BMI) and age (Age).

```{r}
data(NHANES)
# Select relevant columns and drop rows with missing values
nhanes_data <- NHANES %>%
  select(ID, Gender, Age, Race1, Height, Weight, BMI, SmokeNow, PhysActive, PhysActiveDays,
         BPSysAve, BPSysAve, BPDiaAve, TotChol, DirectChol, Diabetes, HealthGen,
         Education, HHIncomeMid, Poverty) %>%
  drop_na() %>% 
  # make it smaller 
  slice_sample(n = 1000) # Randomly select 1000 rows using

# Take a quick look at the data
#paint(nhanes_data)
```

::: 

## [Dataset 1 (NHANES) Variables and their description]{.r-fit-text}

::: {style="font-size: 60%;"}
[[EXCERPT: see complete file in Input Data Folder]]{style="color:#77501a"}

```{r}
#| eval: true
#| output: true
#| echo: false

nhanes_desc <- tribble(
  ~Variable, ~ Type, ~Description,
"X"                ,"int", "xxxx", 
"ID"               ,"int", "xxxxx", 
"SurveyYr"         ,"chr", "yyyy_mm. Ex. 2011_12",    
"Gender"           ,"chr", "Gender (sex) of study participant coded as male or female",
"Age"              ,"int", "##", 
"AgeDecade"        ,"chr", "yy-yy es 20-29",   
# "AgeMonths"        ,"int", "### es 471", 
"Race1"            ,"chr", "Reported race of study participant: Mexican, Hispanic, White, #Black, or Other",
#"Race3"            ,"chr", "Reported race of study participant... Not availale for #2009-10", 

"Education"        ,"chr", "[>= 20 yro]. Ex. 8thGrade, 9-11thGrade, HighSchool, SomeCollege, or CollegeGrad.", 
# "MaritalStatus"    ,"chr", "[>= 20 yro]. Ex. Married, Widowed, Divorced, Separated, # NeverMarried, or LivePartner",  
"HHIncome"         ,"chr","Total annual gross income for the household in US dollars", 
"HHIncomeMid"      ,"int", "Numerical version of HHIncome derived from the middle income # in each category. Ex. 12500 40000" ,   
"Poverty"          ,"dbl", "A ratio of family income to poverty guidelines. Smaller # numbers indicate more poverty Ex.. 0.95 1.74 4.99" , 
# "HomeRooms"        ,"int", "How many rooms are in home of study participant (counting # kitchen but not bath room).", 
# "HomeOwn"          ,"chr", "One of Home, Rent, or Other", 
# "Work"             ,"chr", "NotWorking Working" ,
"Weight"           ,"dbl", "Weight in kg",
# "Length"           ,"lgl", "Recumbent length in cm. Reported for participants aged 0 - 3 years", 
# "HeadCirc"         ,"lgl", "Head circumference in cm. Reported for participants aged 0 years (0 - 6 months)", 
"Height"           ,"dbl",  "Standing height in cm. Reported for participants aged 2 years or older.",
"BMI"              ,"dbl",  "Body mass index (weight/height2 in kg/m2). Reported for participants aged 2 years or older", 
# "BMICatUnder20yrs" ,"lgl",  "Body mass index category. Reported for participants aged 2 to 19 years", 
# "BMI_WHO"          ,"chr",  "Body mass index category. Reported for participants aged 2 years or older", 
"Pulse"            ,"int",  "60 second pulse rate",
"BPSysAve"         ,"int", "Combined systolic blood pressure reading, following the # procedure outlined for BPXSAR",
"BPDiaAve"         ,"int", "Combined diastolic blood pressure reading, following the # procedure outlined for BPXDAR",
#"BPSys1"           ,"int", "Systolic blood pressure in mm Hg – first reading",
#"BPDia1"           ,"int", "Diastolic blood pressure in mm Hg – second reading (consecutive readings)",
#"BPSys2"           ,"int", "Systolic blood pressure in mm Hg – second reading (consecutive readings)",
#"BPDia2"           ,"int", "Diastolic blood pressure in mm Hg – second reading",
#"BPSys3"           ,"int", "Systolic blood pressure in mm Hg third reading (consecutive readings)",
#"BPDia3"           ,"int", "Diastolic blood pressure in mm Hg – third reading (consecutive readings)",
#"Testosterone"     ,"dbl", "Testerone total (ng/dL). Reported for participants aged 6 years or older. Not available for 2009-2010",
#
"DirectChol"       ,"dbl","Direct HDL cholesterol in mmol/L. Reported for participants aged 6 years or older", 
"TotChol"          ,"dbl","Total HDL cholesterol in mmol/L. Reported for participants aged 6 years or older", 
#"UrineVol1"        ,"int", "Urine volume in mL – first test. Reported for participants aged 6 years or older", 
#"UrineFlow1"       ,"dbl", "Urine flow rate in mL/min – first test. Reported for participants aged 6 years or older", 
#"UrineVol2"        ,"int", "Urine volume in mL – second test",
#"UrineFlow2"       ,"dbl","Urine flow rate (urine volume/time since last urination) in mL/min – second test",
"Diabetes"         ,"chr","Study participant told by a doctor or health professional that they have diabetes", 
#"DiabetesAge"      ,"int", "Age of study participant when first told they had diabetes", 
"HealthGen"        ,"chr", "Self-reported rating of health: Excellent, Vgood, Good, Fair, or Poor Fair" , 
#"DaysPhysHlthBad"  ,"int", "Self-reported # of days participant’s physical health was not good out of the past 30 days", 
#"DaysMentHlthBad " ,"int", "Self-reported # of days participant’s mental health was not good out of the past 30 days", 
#"LittleInterest"   ,"chr", "Self-reported # of days where participant had little interest in doing things. Among: None, Several, Majority, #or AlmostAll", 
#"Depressed"        ,"chr", "Self-reported # of days where participant felt down, depressed or hopeless. Among: None, Several, Majority, or #AlmostAll", 
#"nPregnancies"     ,"int", "# times participant has been pregnant",
#"nBabies"          ,"int", "# deliveries resulted in live births", 
#"PregnantNow"      ,"chr", "Pregnancy status ascertained for females 8-59 years of age", 
#
#"Age1stBaby"       ,"int",  "Age of participant at time of first live birth",
#"SleepHrsNight"    ,"int",  "Self-reported # of hours study participant gets at night on weekdays or workdays. For participants aged 16 #years and older", 
#"SleepTrouble"     ,"chr", "Participant [16 years and older] has had trouble sleeping. Coded as Yes or No." ,  
"PhysActive"       ,"chr", "Participant does moderate or vigorous-intensity sports, fitness or recreational activities (Yes or No)." ,  
#"PhysActiveDays"   ,"int", "Number of days in a typical week that participant does moderate or vigorous intensity activity. ", 
#"TVHrsDay"         ,"chr", "Number of hours per day on average participant watched TV over the past 30 days.",
#"CompHrsDay"       ,"chr", "Number of hours per day on average participant used a computer or gaming device over the past 30 day", 
#"TVHrsDayChild"    ,"lgl", "[2-11 yro] Number of hours per day on average participant watched TV over the past 30 days.",
#"CompHrsDayChild"  ,"lgl", "[2-11 yro] Number of hours per day on average participant used a computer or gaming device over the past 30 #day", 
# "Alcohol12PlusYr"  ,"chr",  "Participant has consumed at least 12 drinks of any type of alcoholic beverage in any one year", 
#"AlcoholDay"       ,"int",  "Average number of drinks consumed on days that participant drank alcoholic beverages", 
#"AlcoholYear"      ,"int",  "[>+ 18yro] Estimated number of days over the past year that participant drank alcoholic beverages", 
"SmokeNow"         ,"chr", "Study participant currently smokes cigarettes regularly. (Yes or No)",  
#"Smoke100"         ,"chr", "Study participant has smoked at least 100 cigarettes in their entire life. (Yes pr No)",  
#"Smoke100n"        ,"chr", " Smoker Non-Smoker" , 
#"SmokeAge"         ,"int", "Age study participant first started to smoke cigarettes fairly regularly", 
#"Marijuana"        ,"chr",  "Participant has tried marijuana", 
#"AgeFirstMarij"    ,"int", "Age Participant has tried marijuana first", 
#"RegularMarij"     ,"chr", "Participant has been/is a regular marijuana user (used at least once a month for a year) (Yes or No)",
#"AgeRegMarij"      ,"int", "Age of participant when first started regularly using marijuana", 
#"HardDrugs"        ,"chr", "Participant has tried cocaine, crack cocaine, heroin or methamphetamine (Yes or No)" ,
#"SexEver"          ,"chr", "Participant had had  sex (Yes or No)" ,
#"SexAge"           ,"int", "Age Participant had had sex first time" ,
#"SexNumPartnLife"  ,"int", "Number of opposite sex partners participant has had",
#"SexNumPartYear"   ,"int", "Number of opposite sex partners over the past 12 months",
#"SameSex"          ,"chr", "Participant has had any kind of sex with a same sex partne(Yes or No)" ,
#"SexOrientation"   ,"chr",  "Participant’s sexual orientation One of Heterosexual, Homosexual, Bisexual", 
"..."  ,"...",  "..."#, 
)

kableExtra::kable(nhanes_desc)
```
:::


https://bookdown.org/jbrophy115/bookdown-clinepi/confounding.html

## Causal modeling with CONFOUNDER
+ Z = `Age` = confounder for (X = `SmokeNow`) and blood pressure (Y = `BPSysAve`)
```{r}
data(NHANES)
# Select relevant columns and drop rows with missing values
nhanes_conf <- NHANES %>%
  select(ID, Age, SmokeNow, BPSysAve) %>%
  drop_na()
  
# Take a quick look at the data
paint(nhanes_conf)
```


In this context:

+ `Age` influences both smoking (`SmokeNow`) and blood pressure (`BPSysAve`), making it a *confounder.*
+ `SmokeNow` directly affects `BPSysAve.`

## Causal modeling with CONFOUNDER (viz) 

```{r}
#| output-location: slide 
# linear rel 
qplot (Age, BPSysAve, data = nhanes_conf, color = SmokeNow) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Scatterplot of Age and Blood Pressure by Smoking Status",
       x = "Age",
       y = "Blood Pressure (mmHg)",
       color = "Smoking Status")

```


## --- Regression analysis - unadjusted model
```{r}
# Unadjusted linear model 
model_unadjusted <- lm(BPSysAve ~ SmokeNow, data = nhanes_conf)
summary(model_unadjusted)
 
```

## --- Regression analysis - adjusted model
```{r}

# Adjusted model
model_adjusted <- lm(BPSysAve ~ SmokeNow + Age, data = nhanes_conf)
summary(model_adjusted)
```

## --- Compare models

+ adjusting for age changes the estimated effect of smoking on blood pressure
  + *Unadjusted Model*: This model does not control for age, so any observed relationship between smoking and blood pressure might be confounded.
  + *Adjusted Model*: This model controls for age, providing a more accurate estimate of the causal effect of smoking on blood pressure by accounting for the confounding influence of age.

```{r}
conf_sum <- modelsummary::msummary(list( "NO Confounder" = model_unadjusted, 
                                         "Confounder" = model_adjusted), 
                                   fmt="%.3f", # format
                                   gof_omit = 'DF|Deviance|Log.Lik.|F|R2 Adj.|AIC|BIC',
                                   stars=c('*' = .05, '**' = .01)#,
                                   #coef_rename = cm,
                                   #output = paste(output,"ch19_reg-R.tex",sep="")
) 
```

## --- Compare models

```{r}
#| output: slide
conf_sum
```


## --- Add Predicted Values to the Dataset

```{r}
# Add predicted values from the unadjusted and adjusted models
nhanes_conf <- nhanes_conf %>%
  mutate(
    pred_unadjusted = predict(model_unadjusted),  # Predicted BPSysAve from unadjusted model
    pred_adjusted = predict(model_adjusted)       # Predicted BPSysAve from adjusted model
  )
```


## --- Plot results 1/2

+ **Reshape the Data Using `pivot_longer()`**

We will reshape the dataset so we can plot the predicted values from both models in a faceted plot.

```{r}
# Reshape the data into long format using pivot_longer()
nhanes_long <- nhanes_conf %>%
  pivot_longer(cols = c(pred_unadjusted, pred_adjusted), 
               names_to = "model", values_to = "predicted") %>%
  mutate(model = recode(model, 
                        pred_unadjusted = "Unadjusted: SmokeNow on BPSysAve", 
                        pred_adjusted = "Adjusted: SmokeNow + Age on BPSysAve"))
```

```{r}


```


## --- Plot results 2/2

```{r}
#| output-location: slide 

# Violin plot with dashed line connecting the two conditions
ggplot(nhanes_long, aes(x = factor(SmokeNow), y = BPSysAve, fill = model)) +
  
  # Create the violin plot to show the distribution of blood pressure
  geom_point(color = "grey", alpha = 0.4, position = position_dodge(width = 0.3) ) +
  
  # Overlay the predicted dashed line between the two smoking conditions
  geom_line(aes(y = predicted, group = model, color = model), 
            size = 0.8, linetype = "dashed", position = position_dodge(width = 0.3)) +
  
  # Facet vertically for unadjusted and adjusted models
  facet_wrap(model ~ ., scales = "free_y",ncol = 2) +
  
  # Add labels and title
  labs(title = "Confounder Analysis: Predicted values in Unadjusted vs Adjusted Regression models",
       subtitle = "Age = Confounder",
       x = "Smoking Status",
       y = "Systolic Blood Pressure (BPSysAve)",
       color = "Model",
       fill = "Model"
       ) +
  
  # Customize colors for better contrast
  scale_fill_manual(values = c("Unadjusted: SmokeNow on BPSysAve" ="lightcoral" , 
                               "Adjusted: SmokeNow + Age on BPSysAve" = "lightblue")) +
  scale_color_manual(values = c("Unadjusted: SmokeNow on BPSysAve" = "red", 
                                "Adjusted: SmokeNow + Age on BPSysAve" = "blue")) +
  
  # Minimal theme for clarity
  theme_minimal()+
  theme(legend.position = "none")

```

## Causal modeling with MEDIATOR
<!-- https://tilburgsciencehub.com/topics/analyze/regression/linear-regression/mediation-analysis/ -->


+ M = `BMI` = mediator for the effect of X= `PhysActiveDays` on Y = `BPSysAve`
     
```{r}
data(NHANES)
# Select relevant columns and drop rows with missing values
nhanes_med <- NHANES %>%
  select(Gender, Age, BPSysAve, BMI, PhysActiveDays, Diabetes, SmokeNow) %>%
  drop_na()

# Take a quick look at the data
summary(nhanes_med)
```

## --- Regression analysis - unadjusted (total) model

Before adjusting for any mediator, we fit a simple linear model to estimate the total effect total effect of `BMI` on `Systolic Blood Pressure` without considering `Age`.

+ `Total Effect Model (Unadjusted)` .

```{r}
# Unadjusted model: total effect of smoking on blood pressure
total_effect_model <- lm(BPSysAve ~ PhysActiveDays, data = nhanes_med)
summary(total_effect_model)
```

This model gives the total effect of of `PhysActiveDays` on `Systolic Blood Pressure` (including any *indirect* effects through M = `BMI` or other variables that haven't been included.)


## --- Regression analysis - MEDIATOR model

Now we will set up the mediation models:
 
+ `Mediator model`: This model estimates the effect of physical activity `PhysActiveDays` on BMI `BMI` (= M). This allows us to see whether more physically active days are associated with lower BMI values (which could then affect blood pressure).

```{r}
# Mediator model: SmokeNow affects BMI
mediator_model <- lm(BMI ~ PhysActiveDays, data = nhanes_med)
summary(mediator_model)
```

+ `Outcome Model (Adjusted for BMI)`: This model estimates the effect of `PhysActiveDays` and `BMI` on Systolic Blood Pressure (`BPSysAve`). This is the *adjusted model*, where we adjust for the mediator (`BMI`).

```{r}
# Outcome model: PhysActiveDays and BMI affect BPSysAve
outcome_model <- lm(BPSysAve ~  PhysActiveDays + BMI, data = nhanes_med)
summary(outcome_model)
```


## --- Compare models
```{r}
med_sum <- modelsummary::msummary(list("BMI ~ Mediator Effect" = mediator_model,
                                       "BPSysAve ~ Total Effect" = total_effect_model,
                                       "BPSysAve ~ Outcome" = outcome_model), 
                                  fmt="%.3f", # format
                                  gof_omit = 'DF|Deviance|Log.Lik.|F|R2 Adj.|AIC|BIC',
                                  stars=c('*' = .05, '**' = .01)#,
                                  #coef_rename = cm,
                                  #output = paste(output,"ch19_reg-R.tex",sep="")
) 
```

## --- Compare models
```{r}
#| output: slide
med_sum
```


## --- Calculate the Indirect, Direct, and Total Effects
Now that we have the coefficients from the models, we can calculate the mediation effects manually.

+ **Direct Effect**: This is the coefficient of `PhysActiveDays` in the outcome model, which gives the direct effect of exercise on blood pressure (*controlling* for `BMI` = M)
+ **Indirect Effect**: This is the product of the coefficient from the mediator model (`PhysActiveDays` → `BMI`) and the coefficient from the outcome model (`BMI` → `BPSysAve`).
+ **Total Effect**: This is the coefficient of `PhysActiveDays` in the unadjusted model (*NOT controlling* for `BMI` = M).

```{r}
# Check the names of the coefficients in models...
names(coef(mediator_model))     # [1] BMI ~ PhysActiveDays 
names(coef(outcome_model))      # [1] BPSysAve ~  PhysActiveDays + BMI    
names(coef(total_effect_model)) # [1] BPSysAve ~ PhysActiveDays

# BREAKING DOWN THE ESTIMATE OF PhysActiveDays COEFFICIENT 

# TOTAL effect: The effect of PhysActiveDays on BPSysAve without adjusting for BMI
total_effect <- coef(total_effect_model)["PhysActiveDays"]
total_effect # 0.59

# DIRECT effect: The effect of PhysActiveDays on BPSysAve after adjusting for BMI.
direct_effect <- coef(outcome_model)["PhysActiveDays"]  # Direct effect (controlling for BMI)
direct_effect # 0.621 

# Indirect effect: The portion of the effect on BPSysAve that is mediated through BMI
indirect_effect <- coef(mediator_model)["PhysActiveDays"] * coef(outcome_model)["BMI"]   
indirect_effect # [-0.0821 X 0.382] = -0.0313 
# or 
total_effect - direct_effect # !!!!!!!!  -0.0313
```

## --- Proportion mediated
```{r}
# Proportion mediated: The proportion of the total effect that is mediated through BMI
proportion_mediated <- glue::glue("{round(indirect_effect / total_effect *100, 2)}%")
proportion_mediated #  -0.0531
```
 
## --- Visualise the Indirect, Direct, and Total Effects

"+ Physical Activity" -> "+ Systolic Blood Pressure (BPSysAve)" | "+ BMI -> - BPSysAve "
WTF ??? 

```{r}
# Create a data frame for the effects
effects_df <- data.frame(
  Effect = c("Total Effect", "Direct Effect", "Indirect Effect"),
  Value = c(total_effect, direct_effect, indirect_effect)
)

# Load ggplot2
library(ggplot2)

# Create the bar plot
ggplot(effects_df, aes(x = Effect, y = Value, fill = Effect)) +
  geom_bar(stat = "identity") +
  labs(title = "Mediation Analysis: Total, Direct, and Indirect Effects", 
       subtitle = "Effect (coefficient) of Physical Activity on Systolic Blood Pressure (BPSysAve)",
       y = "", x = "") +
  theme_minimal()
```

## --- Interpret results 

+ The **total effect** tells us the overall relationship between `PhysActiveDays` and `BPSysAve`.

+ The **direct effect** represents the relationship between `PhysActiveDays` and `BPSysAve`, *controlling for* `BMI`.

+ The **indirect effect** (mediated effect) shows how much of the relationship between `PhysActiveDays` and `BPSysAve` is explained through `BMI` (AKA the coef of `PhysActiveDays` in the mediator model times the coef of `BMI` in the outcome model)

+ The **proportion mediated** indicates how much of the total effect is due to the mediation by BMI.


## --- Calculate the Indirect, Direct, and Total Effects (modo2) 
```{r}
library(mediation)
# recall
mediator_model[["call"]][["formula"]]
outcome_model[["call"]][["formula"]]

# Calculate the mediation effect
mediation_model <- mediate(mediator_model, outcome_model, 
                           treat = "PhysActiveDays", 
                           mediator = "BMI", 
                           boot=TRUE, sims=500)
summary(mediation_model)

#                Estimate 95% CI Lower 95% CI Upper p-value  
# ACME            -0.0313      -0.1024         0.04   0.324     = indirect or mediation effect
# ADE              0.6213       0.1479         1.16   0.024 *   = direct effect 
# Total Effect     0.5899       0.1354         1.15   0.020 *   = total effect 
# Prop. Mediated  -0.0531      -0.2969         0.10   0.344     = proportion mediated
```

```{r}
# Add predicted values for BMI, BPSysAve (adjusted), and BPSysAve (total effect)
nhanes_med_pred <- nhanes_med %>%
  mutate(
    pred_bmi = predict(mediator_model),         # Predicted BMI (mediator model)
    pred_bps_adj = predict(outcome_model),      # Predicted BPSysAve (adjusted for Age)
    pred_bps_total = predict(total_effect_model) # Predicted BPSysAve (total effect)
  )
```

## --- Plot results 1/2

Now, let's create a plot that includes:

+ The total effect of BMI on blood pressure (without adjusting for Age).
+ The adjusted effect of BMI on blood pressure (controlling for Age).
+ The mediator effect (BMI's effect on Age).

```{r}
# Check column names
colnames(nhanes_med_pred)

# Reshape the data into long format for faceting
nhanes_long <- nhanes_med_pred %>%
  gather(key = "model", value = "predicted",  pred_bps_adj, pred_bps_total) %>%
  mutate(model = recode(model, 
                        #pred_bmi = "Mediator Effect: BPSysAve on BMI", 
                        pred_bps_adj = "Adjusted Effect: PhysActiveDays + BMI on BPSysAve", 
                        pred_bps_total = "Total Effect: PhysActiveDays on BPSysAve"))
```

## --- Plot results 2/2

```{r}
#| output-location: slide 

# Plot with vertically aligned facets
ggplot(nhanes_long, aes(x = BMI)) +
  geom_point(aes(y = BPSysAve), color = "black", shape = 16, alpha = 0.5) +  # Actual data points for BPSysAve
  geom_line(aes(y = predicted, color = model), size = 1) +  # Predicted values from different models
  facet_grid(model ~ ., scales = "free_y") +  # Facet vertically
  labs(title = "Mediation Analysis: Total, Adjusted, and Mediator Effects",
       x = "BMI",
       y = "Outcome / Mediator",
       color = "Effect Type") +
  theme_minimal()


```

# ----
# QUI
Equitable Equations !!!!!

## [Importing Dataset 2 (Gabor)]{.r-fit-text}

::: {style="font-size: 95%;"}
**Name**:  
**Documentation**:  ..
**Sampling details**:  

```{r}
#| eval: FALSE
# import data
data <- read_csv(paste0(data_in,"food-health.csv")) # 164,848 

# drop observations
dat <-  data %>%
  filter(age >= 30 & age <60) %>% # 7930 
  # new variables: 
  ## Fruit and vegetables per day (grams)
  ## Blood pressure (systolic+diastolic)
  mutate(fv=veggies_n_fruits_gr,
         bp=blood_pressure) %>%
  filter(fv<3200) %>%
  drop_na(bp) %>% # 7358
  # Days per week exercising 
  mutate(exerc=case_when(paq655 <=7 ~ paq655,
                         paq650 ==2 ~ 0)) %>% 
  # Potato chips per day, grams
  mutate(pchips=gr_potato_chips) %>% 
# select variables
  select(id = seqn, age, gender, weight, height, race, 
         fv, exerc, pchips, 
         bp_systolic, bp_diastolic, bp, bmi,  
         smoker, total_cholesterol, hdl,
         hh_income_usd, hh_income_percap, ln_hh_income_percap ,income_cat
                  ) %>% 
  drop_na() # 7358
 

# Hmisc::describe(dat$hh_income)
paint(dat)
```

::: 

## [Dataset 2 (Gabor) Variables and their description]{.r-fit-text}

::: {style="font-size: 60%;"}
[[EXCERPT: see complete file in Input Data Folder]]{style="color:#77501a"}

```{r}
#| eval: FALSE
#| output: true
#| echo: false

food_health_desc <- tribble(
  ~Variable, ~ Type, ~Description,
"X"                ,"int", "xxxx", 
"ID"               ,"int", "xxxxx", 
"SurveyYr"         ,"chr", "yyyy_mm. Ex. 2011_12",    
"Gender"           ,"chr", "Gender (sex) of study participant coded as male or female",
"Age"              ,"int", "##", 
"AgeDecade"        ,"chr", "yy-yy es 20-29",   
# "AgeMonths"        ,"int", "### es 471", 
#"Race1"            ,"chr", "Reported race of study participant: Mexican, Hispanic, White, #Black, or Other",
#"Race3"            ,"chr", "Reported race of study participant... Not availale for #2009-10", 

"Education"        ,"chr", "[>= 20 yro]. Ex. 8thGrade, 9-11thGrade, HighSchool, SomeCollege, or CollegeGrad.", 
# "MaritalStatus"    ,"chr", "[>= 20 yro]. Ex. Married, Widowed, Divorced, Separated, # NeverMarried, or LivePartner",  
# "HHIncome"         ,"chr","Total annual gross income for the household in US dollars", 
# "HHIncomeMid"      ,"int", "Numerical version of HHIncome derived from the middle income # in each category. Ex. 12500 40000" ,   
# "Poverty"          ,"dbl", "A ratio of family income to poverty guidelines. Smaller # numbers indicate more poverty Ex.. 0.95 1.74 4.99" , 
# "HomeRooms"        ,"int", "How many rooms are in home of study participant (counting # kitchen but not bath room).", 
# "HomeOwn"          ,"chr", "One of Home, Rent, or Other", 
# "Work"             ,"chr", "NotWorking Working" ,
  
"Weight"           ,"dbl", "Weight in kg",
# "Length"           ,"lgl", "Recumbent length in cm. Reported for participants aged 0 - 3 years", 
# "HeadCirc"         ,"lgl", "Head circumference in cm. Reported for participants aged 0 years (0 - 6 months)", 
"Height"           ,"dbl",  "Standing height in cm. Reported for participants aged 2 years or older.",
"BMI"              ,"dbl",  "Body mass index (weight/height2 in kg/m2). Reported for participants aged 2 years or older", 
# "BMICatUnder20yrs" ,"lgl",  "Body mass index category. Reported for participants aged 2 to 19 years", 
# "BMI_WHO"          ,"chr",  "Body mass index category. Reported for participants aged 2 years or older", 
"Pulse"            ,"int",  "60 second pulse rate",
# "BPSysAve"         ,"int", "Combined systolic blood pressure reading, following the # procedure outlined for BPXSAR",
"BPDiaAve"         ,"int", "Combined diastolic blood pressure reading, following the # procedure outlined for BPXDAR",
#"BPSys1"           ,"int", "Systolic blood pressure in mm Hg – first reading",
#"BPDia1"           ,"int", "Diastolic blood pressure in mm Hg – second reading (consecutive readings)",
#"BPSys2"           ,"int", "Systolic blood pressure in mm Hg – second reading (consecutive readings)",
#"BPDia2"           ,"int", "Diastolic blood pressure in mm Hg – second reading",
#"BPSys3"           ,"int", "Systolic blood pressure in mm Hg third reading (consecutive readings)",
#"BPDia3"           ,"int", "Diastolic blood pressure in mm Hg – third reading (consecutive readings)",
#"Testosterone"     ,"dbl", "Testerone total (ng/dL). Reported for participants aged 6 years or older. Not available for 2009-2010",
#
"DirectChol"       ,"dbl","Direct HDL cholesterol in mmol/L. Reported for participants aged 6 years or older", 
"TotChol"          ,"dbl","Total HDL cholesterol in mmol/L. Reported for participants aged 6 years or older", 
#"UrineVol1"        ,"int", "Urine volume in mL – first test. Reported for participants aged 6 years or older", 
#"UrineFlow1"       ,"dbl", "Urine flow rate in mL/min – first test. Reported for participants aged 6 years or older", 
#"UrineVol2"        ,"int", "Urine volume in mL – second test",
#"UrineFlow2"       ,"dbl","Urine flow rate (urine volume/time since last urination) in mL/min – second test",
"Diabetes"         ,"chr","Study participant told by a doctor or health professional that they have diabetes", 
"DiabetesAge"      ,"int", "Age of study participant when first told they had diabetes", 
"HealthGen"        ,"chr", "Self-reported rating of health: Excellent, Vgood, Good, Fair, or Poor Fair" , 
#"DaysPhysHlthBad"  ,"int", "Self-reported # of days participant’s physical health was not good out of the past 30 days", 
#"DaysMentHlthBad " ,"int", "Self-reported # of days participant’s mental health was not good out of the past 30 days", 
#"LittleInterest"   ,"chr", "Self-reported # of days where participant had little interest in doing things. Among: None, Several, Majority, #or AlmostAll", 
#"Depressed"        ,"chr", "Self-reported # of days where participant felt down, depressed or hopeless. Among: None, Several, Majority, or #AlmostAll", 
#"nPregnancies"     ,"int", "# times participant has been pregnant",
#"nBabies"          ,"int", "# deliveries resulted in live births", 
#"PregnantNow"      ,"chr", "Pregnancy status ascertained for females 8-59 years of age", 
#
#"Age1stBaby"       ,"int",  "Age of participant at time of first live birth",
#"SleepHrsNight"    ,"int",  "Self-reported # of hours study participant gets at night on weekdays or workdays. For participants aged 16 #years and older", 
#"SleepTrouble"     ,"chr", "Participant [16 years and older] has had trouble sleeping. Coded as Yes or No." ,  
#"PhysActive"       ,"chr", "Participant does moderate or vigorous-intensity sports, fitness or recreational activities (Yes or No)." ,  
#"PhysActiveDays"   ,"int", "Number of days in a typical week that participant does moderate or vigorous intensity activity. ", 
#"TVHrsDay"         ,"chr", "Number of hours per day on average participant watched TV over the past 30 days.",
#"CompHrsDay"       ,"chr", "Number of hours per day on average participant used a computer or gaming device over the past 30 day", 
#"TVHrsDayChild"    ,"lgl", "[2-11 yro] Number of hours per day on average participant watched TV over the past 30 days.",
#"CompHrsDayChild"  ,"lgl", "[2-11 yro] Number of hours per day on average participant used a computer or gaming device over the past 30 #day", 
# "Alcohol12PlusYr"  ,"chr",  "Participant has consumed at least 12 drinks of any type of alcoholic beverage in any one year", 
#"AlcoholDay"       ,"int",  "Average number of drinks consumed on days that participant drank alcoholic beverages", 
#"AlcoholYear"      ,"int",  "[>+ 18yro] Estimated number of days over the past year that participant drank alcoholic beverages", 
#"SmokeNow"         ,"chr", "Study participant currently smokes cigarettes regularly. (Yes or No)",  
#"Smoke100"         ,"chr", "Study participant has smoked at least 100 cigarettes in their entire life. (Yes pr No)",  
#"Smoke100n"        ,"chr", " Smoker Non-Smoker" , 
#"SmokeAge"         ,"int", "Age study participant first started to smoke cigarettes fairly regularly", 
#"Marijuana"        ,"chr",  "Participant has tried marijuana", 
#"AgeFirstMarij"    ,"int", "Age Participant has tried marijuana first", 
#"RegularMarij"     ,"chr", "Participant has been/is a regular marijuana user (used at least once a month for a year) (Yes or No)",
#"AgeRegMarij"      ,"int", "Age of participant when first started regularly using marijuana", 
#"HardDrugs"        ,"chr", "Participant has tried cocaine, crack cocaine, heroin or methamphetamine (Yes or No)" ,
#"SexEver"          ,"chr", "Participant had had  sex (Yes or No)" ,
#"SexAge"           ,"int", "Age Participant had had sex first time" ,
#"SexNumPartnLife"  ,"int", "Number of opposite sex partners participant has had",
#"SexNumPartYear"   ,"int", "Number of opposite sex partners over the past 12 months",
#"SameSex"          ,"chr", "Participant has had any kind of sex with a same sex partne(Yes or No)" ,
#"SexOrientation"   ,"chr",  "Participant’s sexual orientation One of Heterosexual, Homosexual, Bisexual", 
"..."  ,"...",  "..."#, 
)

kableExtra::kable(nhanes_desc)
```
:::


https://bookdown.org/jbrophy115/bookdown-clinepi/confounding.html

## Causal modeling with CONFOUNDER
```{r}
#| eval: FALSE
# Select relevant columns and drop rows with missing values
dat_conf <- dat %>%
  select(id, age, bp_systolic, fv, exerc, bmi, smoker, gender, hh_income_usd, ln_hh_income_percap ) %>%
  drop_na() %>% 
  filter(fv <=1500)

# Take a quick look at the data
summary(dat_conf)

Hmisc::describe(dat_conf$bp_systolic) # Y
Hmisc::describe(dat_conf$bmi) # X
Hmisc::describe(dat_conf$age) # Z (confounder )
Hmisc::describe(dat_conf$exerc)  # Z (confounder )
Hmisc::describe(dat_conf$ln_hh_income_percap) # Z (confounder )


```

In this context:

+ `exerc` influences both smoking (`fv`) and blood pressure (`bp_systolic`), making it a confounder (albeit as PROXY for socio-economic status)
+ `fv` directly affects `bp_systolic`

## --- Regression analysis  
```{r}
#| eval: FALSE
# Unadjusted model
model_unadjusted <- lm( bp_systolic ~ bmi + ln_hh_income_percap , data = dat_conf)
summary(model_unadjusted)

# Adjusted model
model_adjusted <- lm(bp_systolic ~ bmi + age + ln_hh_income_percap, data = dat_conf)
summary(model_adjusted)

```

+ adjusting for Z changes the estimated effect of X on Y
  + *Unadjusted Model*: This model does not control for Z, which is a confounder of the relationship between X and Y. 
  + *Adjusted Model*: This model controls for Z, providing a more accurate estimate of the causal effect of X on Y.
 
## --- Add Predicted Values to the Dataset

```{r}
#| eval: FALSE
model_unadjusted$coefficients 
model_adjusted$coefficients

# Get predicted values using augment and attach to the original dataset
dat_unadjusted <- broom::augment(model_unadjusted, data = dat_conf) %>%
  select( -.resid,  -.hat, -.sigma, -.cooksd, -.std.resid ) %>%
  rename(pred_unadjusted = .fitted)

dat_adjusted <- broom::augment(model_adjusted, data = dat_conf) %>%
  select( -.resid,     -.hat, -.sigma, -.cooksd, -.std.resid ) %>%
  rename(pred_adjusted = .fitted)

# Combine the predicted values from both models
dat_conf_long <- dat_unadjusted %>%
  left_join(dat_adjusted, by = "id", suffix=c("",".y"))%>%
  select(-ends_with(".y")) %>% 
  # Convert the data from wide to long format
  pivot_longer(cols = c(pred_unadjusted, pred_adjusted), 
               names_to = "model", 
               values_to = "predicted")

# Check the structure of the long dataset
str(dat_conf_long)

```


## --- Plot results 1/2

+ **Reshape the Data Using `pivot_longer()`**

We will reshape the dataset so we can plot the predicted values from both models in a faceted plot.

```{r}
#| eval: FALSE
# check 
dat_conf_long %>% group_by(model) %>% 
  summarise(mean(predicted), sd(predicted))

```
 

## --- Plot results 2/2

```{r}
#| eval: FALSE
#| output-location: slide 

ggplot(dat_conf, aes(x = bmi, y = bp_systolic)) +
  geom_point(color = "grey", alpha = 1) +  
  geom_point(data = dat_unadjusted, aes(x = bmi, y = pred_unadjusted), 
             color = "#e60066", alpha = 0.20) +
  geom_point(data = dat_adjusted, aes(x = bmi, y = pred_adjusted), 
             color = "#399B23", alpha = 0.20) +
  # Add labels and title
  labs(title = "Confounder Analysis: Unadjusted (pink) vs Adjusted (green) fitted values",
       x = "BMI",
       y = "Systolic Blood Pressure (BPSysAve)"
       )  +
    # Minimal theme for clarity
  theme_minimal() +
  theme(legend.position = "none")

```



## --- Different 
https://nicholasrjenkins.science/post/data_viz_r/data_visualization_r/

## --- Regression analysis  
```{r}
#| eval: FALSE
# Unadjusted model
model_unadjusted <- lm( bp_systolic ~ bmi  , data = dat_conf)  
 # Adjusted model
model_adjusted <- lm(bp_systolic ~ bmi + age  , data = dat_conf)  
```

 
```{r}
#| eval: FALSE
tidy(model_unadjusted, conf.int = TRUE ) %>% 
  ggplot(data = ., 
         mapping = aes(x = estimate, y = term, xmin = conf.low, xmax = conf.high)) +
  geom_pointrange() +
  labs(title = "Model Estimates of bmi on bp_systolic",
       x = "Coefficient Estimate",
       y = "Predictor",
       caption = "Models fit with OLS. Error bars show the 95% confidence interval.") +
  scale_y_discrete(labels = c("Intercept",  "bmi")) +
  ggpubr::theme_pubclean(flip = TRUE)
 
``` 
 
```{r}
#| eval: FALSE
tidy(model_adjusted, conf.int = TRUE ) %>% 
  ggplot(data = ., 
         mapping = aes(x = estimate, y = term, xmin = conf.low, xmax = conf.high)) +
  geom_pointrange() +
  labs(title = "Model Estimates of bmi + age on bp_systolic",
       x = "Coefficient Estimate",
       y = "Predictor",
       caption = "Models fit with OLS. Error bars show the 95% confidence interval.") +
  scale_y_discrete(labels = c("Intercept", "age", "bmi")) +
  ggpubr::theme_pubclean(flip = TRUE)
 
```

 
```{r}
#| eval: FALSE
fit_unadj_results <- tidy(model_unadjusted, conf.int = TRUE) %>% 
  mutate(model = "Model Unadjusted")

fit_adj_results <- tidy(model_adjusted, conf.int = TRUE) %>% 
  mutate(model = "Model Adjusted")

model_results <- bind_rows(fit_unadj_results, fit_adj_results)
```
 
```{r}
#| eval: FALSE
ggplot(data = model_results, 
       aes(x = estimate, y = term, xmin = conf.low, xmax = conf.high, 
           color = model, shape = model)) +
  geom_pointrange(position = position_dodge(width = 0.5)) +
  labs(title = "Model Estimates of BMI and age on REM Sleep",
       x = "Coefficient Estimate",
       y = "Predictor",
       caption = "Models fit with OLS. Error bars show the 95% confidence interval.",
        color = "Model:",
       shape = "Model:") +
  scale_y_discrete(labels = c("Intercept", "age", "bmi")) +
  ggpubr::theme_pubclean(flip = FALSE)

```
 
## Final thoughts

::: {style="font-size: 95%;"}
::: {style="color:#77501a"}


+ .....

:::

:::



